---
title: "VIDA: Causes of death for children"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      message=FALSE,
                      warning=FALSE)

library(tidyverse)
library(flextable)
library(knitr)
library(kableExtra)
source("kalter_child.R")
# source("liu_child.R")
```

# EAVA algorithm 

Expert Algorithm for Verbal Autopsy (EAVA) based on

> Kalter et al., [Direct estimates of national neonatal and child cause–specific mortality proportions in Niger by expert algorithm and physician–coded analysis of verbal autopsy interviews](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4416334/), J Glob Health. 2015 Jun; 5(1): 010415

and

> Liu et al., [Deriving causes of child mortality by re–analyzing national verbal autopsy data applying a standardized computer algorithm in Uganda, Rwanda and Ghana](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4467513/), J Glob Health. 2015 Jun; 5(1): 010414

```{r}

# causes is a list created by get_causes ; 
# 1st entry is primary cause, others are possible comborbidities
clean_causes <- function( causes ){
  primary_causes <- sapply(causes,"[[",1) 
  comorbidities <- character( length = length( primary_causes ) ) 
  for( i in 1:length( causes ) ){
    if( length(causes[[i]]) == 1 ){
      comorbidities[i] <- ""
    } else{
      comorbidities[i] <- paste0( causes[[i]][-1], collapse=", " )
    }
  }
  return( data.frame( Primary = primary_causes, Comorbidities = comorbidities) )
}

tabulate_causes <- function( cleaned_causes ){

tallies <- cleaned_causes %>%
  group_by( Primary ) %>%
  tally( name="Count") 

comorbid <-  cleaned_causes %>% 
  group_by(Primary) %>% 
  summarize( Comorbidities = paste0( unique( unlist( strsplit( paste0(Comorbidities, collapse = ", "), ", ", fixed=TRUE))), collapse = ", "))

return( tallies %>% left_join( comorbid, by="Primary" ) %>% arrange( desc(Count) ))
}

```

# Results for Mali data

```{r}
df <- read_csv("data/mali_babel/CHAMPS_Verbal_Autopsy_Form_results_babel.csv")
causes <- get_causes(df, format="champs2")
causes <- tabulate_causes( clean_causes( causes ) )

causes %>%
  rename( "Possible comorbidities" = Comorbidities) %>%
  regulartable() %>% 
  theme_zebra() %>% 
  set_caption("CHAMPS_Verbal_Autopsy_Form_results.csv") %>% 
  autofit()

# regulartable( cbind( death_dates(df, format="champs2"), interview_dates(df, format="champs2") ) ) %>% 
#   autofit()

```

```{r}
df <- read_csv("data/mali_babel/CHAMPS_Verbal_Autopsy_Form_v2_04_1_results_babel.csv")
causes <- get_causes( df, format="champs2")
causes <- data.frame( Cause = causes )
causes %>% 
  group_by(Cause) %>% 
  tally %>% 
  regulartable() %>% 
  theme_zebra() %>% 
  set_caption("CHAMPS_Verbal_Autopsy_Form_v2_04_1_results.csv") %>% 
  autofit()

# regulartable( cbind( death_dates(df, format="champs2"), interview_dates(df, format="champs2") ) ) %>% 
#   autofit()

```

```{r}
df <- read_csv("data/mali_babel/CHAMPS_Verbal_Autopsy_Form_v2_04_results_babel.csv")
causes <- get_causes( df, format="champs2")
causes <- data.frame( Cause = causes )
causes %>% 
  group_by(Cause) %>% 
  tally %>% 
  regulartable() %>% 
  theme_zebra() %>% 
  set_caption("CHAMPS_Verbal_Autopsy_Form_v2_04_results.csv") %>% 
  autofit()

# regulartable( cbind( death_dates(df, format="champs2"), interview_dates(df, format="champs2") ) ) %>% 
#   autofit()

```

```{r}
df <- read_csv("data/mali_babel/VA_Death_of_a_child_under_4_weeks_to_59_months_results_21MAR2018_babel.csv")
causes <- get_causes( df, format="champs1")
causes <- data.frame( Cause = causes )
causes %>% 
  group_by(Cause) %>% 
  tally %>% 
  regulartable() %>% 
  theme_zebra() %>% 
  set_caption("VA_Death_of_a_child_under_4_weeks_to_59_months_results_21MAR2018.csv") %>% 
  autofit()

# regulartable( cbind( death_dates(df, format="champs1"), interview_dates(df, format="champs1") ) ) %>% 
#   autofit()

```

# Results for Gambia data

## BN HDSS 

```{r}
df <- read_csv("data/gambia_babel/bn_hdss/VA_Child_WHO2016_babel.csv")
causes <- get_causes( df, format="c16")
causes <- data.frame( Cause = causes )
causes %>% 
  group_by(Cause) %>% 
  tally %>% 
  regulartable() %>% 
  theme_zebra() %>% 
  set_caption("VA_Child_WHO2016.csv") %>% 
  autofit()

# regulartable( cbind( death_dates(df, format="c16"), interview_dates(df, format="c16") ) ) %>% 
#   autofit()

```

```{r}
df <- read_csv("data/gambia_babel/bn_hdss/VA_ChildForm_WHO2012_babel.csv")
causes <- get_causes( df, format="c12_16")
causes <- data.frame( Cause = causes )
causes %>% 
  group_by(Cause) %>% 
  tally %>% 
  regulartable() %>% 
  theme_zebra() %>% 
  set_caption("VA_ChildForm_WHO2012.csv") %>% 
  autofit()

# regulartable( cbind( death_dates(df, format="c12_16"), interview_dates(df, format="c12_16") ) ) %>% 
#   autofit()

```

## BS HDSS 

```{r}
df <- read_csv("data/gambia_babel/bs_hdss/VA_Child_Indepth_2008-2012_clean duration_babel.csv")
causes <- get_causes( df, format="c08_12")
causes <- data.frame( Cause = causes )
causes %>% 
  group_by(Cause) %>% 
  tally %>% 
  regulartable() %>% 
  theme_zebra() %>% 
  set_caption("VA_Child_Indepth_2008-2012_clean duration.csv") %>% 
  autofit()

```


```{r}
df <- read_csv("data/gambia_babel/bs_hdss/VA_Child_Indepth_2008-2012_babel.csv")
causes <- get_causes( df, format="c08_12")
causes <- data.frame( Cause = causes )
causes %>% 
  group_by(Cause) %>% 
  tally %>% 
  regulartable() %>% 
  theme_zebra() %>% 
  set_caption("VA_Child_Indepth_2008-2012.csv") %>% 
  autofit()

```


```{r}
df <- read_csv("data/gambia_babel/bs_hdss/VA_Child_WHO2012-mid 2016_babel.csv")
causes <- get_causes( df, format="c12_16")
causes <- data.frame( Cause = causes )
causes %>% 
  group_by(Cause) %>% 
  tally %>% 
  regulartable() %>% 
  theme_zebra() %>% 
  set_caption("VA_Child_WHO2012-mid 2016.csv") %>% 
  autofit()

# regulartable( cbind( death_dates(df, format="c12_16"), interview_dates(df, format="c12_16") ) ) %>% 
#   autofit()

```


```{r}
df <- read_csv("data/gambia_babel/bs_hdss/VA_Child_WHO2016_babel.csv")
causes <- get_causes( df, format="c16")
causes <- data.frame( Cause = causes )
causes %>% 
  group_by(Cause) %>% 
  tally %>% 
  regulartable() %>% 
  theme_zebra() %>% 
  set_caption("VA_Child_WHO2016.csv") %>% 
  autofit()

# regulartable( cbind( death_dates(df, format="c16"), interview_dates(df, format="c16") ) ) %>% 
#   autofit()

```

