---
title: "VIDA: Causes of death for children"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(flextable)
source("kalter_child.R")
```

## EAVA algorithm 

Expert Algorithm for Verbal Autopsy (EAVA) based on

> Kalter et al., [Direct estimates of national neonatal and child cause–specific mortality proportions in Niger by expert algorithm and physician–coded analysis of verbal autopsy interviews](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4416334/), J Glob Health. 2015 Jun; 5(1): 010415

## Sites and available data

* Kenya
  * ?? datasets
  * time range
* The Gambia 
  * ?? datasets
  * time range
* Mali
  * ?? datasets
  * time range

## Results

```{r echo=FALSE, message=FALSE}
df <- read_csv("data/gambia_babel/bn_hdss/VA_Child_WHO2016_babel.csv")
causes <- get_causes( df, format="c16")
causes <- data.frame( Cause = causes )
causes %>% group_by(Cause) %>% tally %>% regulartable() %>% autofit()
```




