---
title: "VA Diarrhea Statistics"
author: "VIDA Project"
date: "`r Sys.time()`"
output:
  html_document:
    theme: yeti
    toc: yes
    toc_float: yes
    toc_depth: 3
  pdf_document: default
---

```{r setup, include=FALSE, echo = FALSE}
# Basic knitr options
library(knitr)
opts_chunk$set(comment = NA, 
               echo = FALSE,
               warning = FALSE, 
               message = FALSE, 
               error = TRUE, 
               cache = FALSE)
options(scipen = '999')
```

```{r}

# Load packages and WHO VA instrument
library(tidyverse)
library(stringi)
library(readxl)
library(pander)
library(gridExtra)
library(descr)

WHOVA2016 <- read_excel("docs/WHOVA2016-Instrument_Manual_and_Questionnaire/WHOVA2016_v1_5_3_XLS_form_for_ODK.xlsx")

```

# Relevant variables

In the 2016 WHO and CHAMPS VA surveys, variables `Id10181` through `Id10187` are most directly relevant:

```{r}
stool_vars <- c('Id10181', 'Id10182', 'Id10183', 'Id10184', 'Id10185', 'Id10186', 'Id10187')
idx <- which( WHOVA2016$name %in% stool_vars )
knitr::kable( data.frame(Question = WHOVA2016$`label::English`[idx] ))

```

In the 2012 WHO VA survey, variables `3B280` through `3B300` are most relevant:

* `3B280` : Did (s)he have diarrhoea?
* `3B290` : For how long did (s)he have diarrhoea? (time in weeks)
* `3B290_choice` : acute (< 2 weeks), persistent (2--4 weeks), recurrent ( >= 4 weeks)
* `3B300` : At any time during the final illness was there blood in the stools?

# Gambia child data

```{r}

# get the 2016 data:
####################

gambia_bn <- read_csv("va-data/VA_Data_Gambia_Site_31Oct2017/BN_HDSS/csv/VA_Child_WHO2016.csv")
names(gambia_bn) <- gsub( "child", "Id", names( gambia_bn ) )

gambia_bn <- gambia_bn %>% select(Id10181:Id10187)
gambia_bn <- gambia_bn %>%
  mutate( Id10181 = recode_factor(.x = Id10181,
                                  "1" = "Yes",
                                  "2" = "No" ,
                                  "3" = "Don't know")) %>%
  mutate( Id10185 = recode_factor(.x = Id10185, 
                                  "1" = "Yes",
                                  "2" = "No",
                                  "3" = "Don't know")) %>%
  mutate( Country = "Gambia", Site = "BN_HDSS", Category = "Child", Year = "2016")

gambia_bs <- read_csv("va-data/VA_Data_Gambia_Site_31Oct2017/BS_HDSS/csv/VA_Child_WHO2016.csv")
names(gambia_bs) <- gsub( "child", "Id", names( gambia_bs ) )

gambia_bs <- gambia_bs %>% select(Id10181:Id10187)
gambia_bs <- gambia_bs %>%
  mutate( Id10181 = recode_factor(.x = Id10181,
                                  "1" = "Yes",
                                  "2" = "No" ,
                                  "3" = "Don't know")) %>%
  mutate( Id10185 = recode_factor(.x = Id10185, 
                                  "1" = "Yes",
                                  "2" = "No",
                                  "3" = "Don't know")) %>%
  mutate( Country = "Gambia", Site = "BS_HDSS", Category = "Child", Year = "2016")

# combine the 2016 data:
########################

gambia_2016 <- rbind( gambia_bn, gambia_bs)

# now get the 2012 data :
#########################

gambia_bn <- read_csv("va-data/VA_Data_Gambia_Site_31Oct2017/BN_HDSS/csv/VA_ChildForm_WHO2012.csv")
gambia_bn <- gambia_bn %>% select(Q3B280:Q3B300)
gambia_bn <- gambia_bn %>%
  mutate( Q3B280 = recode_factor(.x = Q3B280,
                                  "1" = "Yes",
                                  "2" = "No" ,
                                  "3" = "Don't know")) %>%
  mutate( Country = "Gambia", Site = "BN_HDSS", Category = "Child", Year = "2012")

gambia_bs <- read_csv("va-data/VA_Data_Gambia_Site_31Oct2017/BS_HDSS/csv/VA_Child_WHO2012-mid 2016.csv")
gambia_bs <- gambia_bs %>% select(Q3B280:Q3B300)
gambia_bs <- gambia_bs %>%
  mutate( Q3B280 = recode_factor(.x = Q3B280,
                                  "1" = "Yes",
                                  "2" = "No" ,
                                  "3" = "Don't know")) %>%
  mutate( Country = "Gambia", Site = "BS_HDSS", Category = "Child", Year = "2012 to mid-2016")

# combine the 2012 to mid-2016 data :
#####################################

gambia_2012_2016 <- rbind( gambia_bn, gambia_bs)

```

## Basic incidence

### 2016 data: `Id10181`

```{r}

addmargins( table( gambia_2016$Id10181, gambia_2016$Site ) )

prop.table( table( gambia_2016$Id10181, gambia_2016$Site ), 2)

bn <- gambia_2016 %>% 
  filter( Site == "BN_HDSS" ) %>% 
  drop_na( Id10181 ) %>% 
  ggplot( aes( x= Id10181, 
               y = 100*( (..count..) / sum(..count..) ))) + 
  geom_bar() + 
  labs( title = "BN HDSS", y = "Percentage" ) + ylim(0,100)

bs <- gambia_2016 %>% 
  filter( Site == "BS_HDSS" ) %>% 
  drop_na( Id10181 ) %>% 
  ggplot( aes( x= Id10181, 
               y = 100*( (..count..) / sum(..count..) ))) + 
  geom_bar() + 
  labs( title = "BS HDSS", y = "Percentage" ) + ylim(0,100)

grid.arrange( bn, bs, ncol=2 )

```

### 2012 to mid-2016 data: `Q3B280`

```{r}

addmargins( table( gambia_2012_2016$Q3B280, gambia_2012_2016$Site ) )

prop.table( table( gambia_2012_2016$Q3B280, gambia_2012_2016$Site ), 2 )

bn <- gambia_2012_2016 %>% 
  filter( Site == "BN_HDSS" ) %>% 
  drop_na( Q3B280 ) %>% 
  ggplot( aes( x= Q3B280, 
               y = 100*( (..count..) / sum(..count..) ))) + 
  geom_bar() + 
  labs( title = "BN HDSS", y = "Percentage" ) + ylim(0,100)

bs <- gambia_2012_2016 %>% 
  filter( Site == "BS_HDSS" ) %>% 
  drop_na( Q3B280 ) %>% 
  ggplot( aes( x= Q3B280, 
               y = 100*( (..count..) / sum(..count..) ))) + 
  geom_bar() + 
  labs( title = "BS HDSS", y = "Percentage" ) + ylim(0,100)

grid.arrange( bn, bs, ncol=2 )

```

## Loose/liquid stool until death 

### 2016 data: `Id10185`

```{r}

addmargins( table( gambia_2016$Id10185, gambia_2016$Site ) )

prop.table( table( gambia_2016$Id10185, gambia_2016$Site ), 2 )

bn <- gambia_2016 %>% 
  filter( Site == "BN_HDSS" ) %>% 
  drop_na( Id10185 ) %>% 
  ggplot( aes( x= Id10185, y = 100*( (..count..) / sum(..count..) ) )) + 
  geom_bar() + 
  labs( title = "BN HDSS", y = "Percentage" ) + ylim(0,100)

bs <- gambia_2016 %>% 
  filter( Site == "BS_HDSS" ) %>% 
  drop_na( Id10185 ) %>% 
  ggplot( aes( x= Id10185, y = 100*( (..count..) / sum(..count..) ) )) + 
  geom_bar() + 
  labs( title = "BS HDSS", y = "Percentage" ) + ylim(0,100)

grid.arrange( bn, bs, ncol=2 )

```

# Gambia neonatal data

## 2016 data 

```{r}

# get the 2016 data:
####################

gambia_bn <- read_csv("va-data/VA_Data_Gambia_Site_31Oct2017/BN_HDSS/csv/VA_Neonates_WHO2016.csv")
names(gambia_bn) <- gsub( "neonate", "Id", names( gambia_bn ) )

gambia_bn <- gambia_bn %>% select(Id10181:Id10186)

# gambia_bn <- gambia_bn %>%
#   mutate( Id10181 = recode_factor(.x = Id10181,
#                                   "1" = "Yes",
#                                   "2" = "No" ,
#                                   "3" = "Don't know")) %>%
#   mutate( Country = "Gambia", Site = "BN_HDSS", Category = "Child", Year = "2016")

gambia_bs <- read_csv("va-data/VA_Data_Gambia_Site_31Oct2017/BS_HDSS/csv/VA_Neonates_WHO2016.csv")
names(gambia_bs) <- gsub( "neonate", "Id", names( gambia_bs ) )

gambia_bs <- gambia_bs %>% select(Id10181:Id10186)

gambia_bs <- gambia_bs %>%
  mutate( Id10181 = recode_factor(.x = Id10181,
                                  "1" = "Yes",
                                  "2" = "No" ,
                                  "3" = "Don't know")) %>%
  mutate( Country = "Gambia", Site = "BS_HDSS", Category = "Child", Year = "2016")

```

There is no data for the most directly relevant variables from `BN_HDSS` site:

```{r}
knitr::kable( gambia_bn )
```

And not much from the `BS_HDSS` site:

```{r}

table( Id10181 = gambia_bs$Id10181, useNA="ifany")

```

## 2012 to mid-2016 data

```{r}

# get the 2012 to mid-2016 data:
################################

gambia_bn <- read_csv("va-data/VA_Data_Gambia_Site_31Oct2017/BN_HDSS/csv/VA_Neonates_WHO2012.csv")
names(gambia_bn) <- gsub( "neonate", "Id", names( gambia_bn ) )

gambia_bn <- gambia_bn %>% select(Q3B280:Q3B300)
gambia_bn <- gambia_bn %>%
  mutate( Q3B280 = recode_factor(.x = Q3B280,
                                  "1" = "Yes",
                                  "2" = "No" ,
                                  "3" = "Don't know")) %>%
  mutate( Country = "Gambia", Site = "BN_HDSS", Category = "Child", Year = "2012 to mid-2016")

gambia_bs <- read_csv("va-data/VA_Data_Gambia_Site_31Oct2017/BS_HDSS/csv/VA_Neonates_WHO2012-mid-2016.csv")
names(gambia_bs) <- gsub( "neonate", "Id", names( gambia_bs ) )

gambia_bs <- gambia_bs %>% select(Q3B280:Q3B300)
gambia_bs <- gambia_bs %>%
  mutate( Q3B280 = recode_factor(.x = Q3B280,
                                  "1" = "Yes",
                                  "2" = "No" ,
                                  "3" = "Don't know")) %>%
  mutate( Country = "Gambia", Site = "BS_HDSS", Category = "Child", Year = "2012 to mid-2016")

# combine the 2012 to mid-2016 data :
#####################################

gambia_2012_2016 <- rbind( gambia_bn, gambia_bs)

```

### 2012 to mid-2016 data: `Q3B280`

```{r}

addmargins( table( gambia_2012_2016$Q3B280, gambia_2012_2016$Site ) )

prop.table( table( gambia_2012_2016$Q3B280, gambia_2012_2016$Site ), 2 )

bn <- gambia_2012_2016 %>% 
  filter( Site == "BN_HDSS" ) %>% 
  drop_na( Q3B280 ) %>% 
  ggplot( aes( x= Q3B280, 
               y = 100*( (..count..) / sum(..count..) ))) + 
  geom_bar() + 
  labs( title = "BN HDSS", y = "Percentage" ) + ylim(0,100)

bs <- gambia_2012_2016 %>% 
  filter( Site == "BS_HDSS" ) %>% 
  drop_na( Q3B280 ) %>% 
  ggplot( aes( x= Q3B280, 
               y = 100*( (..count..) / sum(..count..) ))) + 
  geom_bar() + 
  labs( title = "BS HDSS", y = "Percentage" ) + ylim(0,100)

grid.arrange( bn, bs, ncol=2 )

```

# Mali CHAMPS 2017 data

## Basic incidence: `Id10181`

```{r}

# get the 2017 data:
####################

mali_champs <- read_csv("va-data/Mali VA/WHO 2017/CHAMPS_Verbal_Autopsy_Form_results.csv")
idx <- grep( "Id1018", names( mali_champs ))
mali_champs <- mali_champs[,idx]
names( mali_champs ) <- c("Id10181", "Id10182", "Id10183", "Id10184", "Id10185", "Id10186", "Id10187", "Id10188", "Id10189")

mali_champs <- mali_champs %>% select(Id10181:Id10187)

mali_champs_v2_04 <- read_csv("va-data/Mali VA/WHO 2017/CHAMPS_Verbal_Autopsy_Form_v2_04_results.csv")
idx <- grep( "Id1018", names( mali_champs_v2_04 ))
mali_champs_v2_04 <- mali_champs_v2_04[,idx]
names( mali_champs_v2_04 ) <- c("Id10181", "Id10182", "Id10183", "Id10184", "Id10185", "Id10186", "Id10187", "Id10188", "Id10189")

mali_champs_v2_04 <- mali_champs_v2_04 %>% select(Id10181:Id10187)

mali_champs_v2_04_1 <- read_csv("va-data/Mali VA/WHO 2017/CHAMPS_Verbal_Autopsy_Form_v2_04_1_results.csv")
idx <- grep( "Id1018", names( mali_champs_v2_04_1 ))
mali_champs_v2_04_1 <- mali_champs_v2_04_1[,idx]
names( mali_champs_v2_04_1 ) <- c("Id10181", "Id10182", "Id10183", "Id10184", "Id10185", "Id10186", "Id10187", "Id10188", "Id10189")

mali_champs_v2_04_1 <- mali_champs_v2_04_1 %>% select(Id10181:Id10187)

```

### `CHAMPS_Verbal_Autopsy_Form_results.csv`

```{r}

addmargins( table( mali_champs$Id10181 ) )

prop.table( table( mali_champs$Id10181 ) )

mali_champs %>% 
  drop_na( Id10181 ) %>% 
  ggplot( aes( x= Id10181, 
               y = 100*( (..count..) / sum(..count..) ))) + 
  geom_bar() + 
  labs( title = "Mali CHAMPS 2017", y = "Percentage" ) + ylim(0,100)

```

### `CHAMPS_Verbal_Autopsy_Form_v2_04_results.csv`

```{r}

addmargins( table( mali_champs_v2_04$Id10181 ) )

prop.table( table( mali_champs_v2_04$Id10181 ) )

mali_champs_v2_04 %>% 
  drop_na( Id10181 ) %>% 
  ggplot( aes( x= Id10181, 
               y = 100*( (..count..) / sum(..count..) ))) + 
  geom_bar() + 
  labs( title = "Mali CHAMPS 2017 v2_04", y = "Percentage" ) + ylim(0,100)

```

### `CHAMPS_Verbal_Autopsy_Form_v2_04_1_results.csv`

```{r}

addmargins( table( mali_champs_v2_04_1$Id10181 ) )

prop.table( table( mali_champs_v2_04_1$Id10181 ) )

mali_champs_v2_04_1 %>% 
  drop_na( Id10181 ) %>% 
  ggplot( aes( x= Id10181, 
               y = 100*( (..count..) / sum(..count..) ))) + 
  geom_bar() + 
  labs( title = "Mali CHAMPS 2017 v2_04_1", y = "Percentage" ) + ylim(0,100)

```