---
title: "Expert Algorithms for Verbal Autopsy and the VIDA project"
author: "DataBrew LLC"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    toc: yes
    toc_depth: 3
  pdf_document:
    toc: yes
    toc_depth: '3'
  word_document:
    toc: yes
    toc_depth: '3'
bibliography: eava.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      message = FALSE,
                      warning = FALSE)
library(tidyverse)
library(flextable)
library(knitr)
library(kableExtra)
source("eava.R")
```

# Introduction

## Expert Algorithms for Verbal Autopsy 

To determine causes of death from available verbal autopsy (VA) data, we have implemented several Expert Algorithms for Verbal
Autopsy (EAVA), following the details provided in the supplements to @kalter and @liu. As discussed in these papers,
there are separate algorithms for neonatal decedents (infants under 1 month old at death) and for child decedents (older than 
1 month at time of death). In each case, the algorithms proceed through a hierarchy of conditions to determine the cause of death;
the first condition whose criteria are satisfied is determined to be the primary cause of death, and the remaining conditions 
whose criteria are satisfied are identified as possible comorbidities. The specific hierarchies for the algorithms implemented 
thus far are as follows:

### Kalter hierarchy for neonates

1. Neonatal tetanus 
2. Congenital malformation 
3. Birth asphyxia 
4. Birth injury 
5. Preterm delivery with respiratory distress syndrome 
6. Meningitis 
7. Diarrhea 
8. Pneumonia 
9. Possible diarrhea 
10. Possible pneumonia 
11. Sepsis 
12. Neonatal jaundice 
13. Neonatal hemorrhagic syndrome 
14. Sudden unexplained death 
15. Preterm delivery 
16. Unspecified 

### Kalter hierarchy for children

1. Injury 
2. AIDS 
3. Malnutrition (underlying) 
4. Measles 
5. Meningitis 
6. Dysentery 
7. Diarrhea 
8. Pertussis 
9. Pneumonia 
10. Malaria 
11. Possible dysentery 
12. Possible diarrhea 
13. Possible pneumonia 
14. Hemorrhagic fever 
15. Other infection 
16. Possible malaria 
17. Malnutrition 
18. Unspecified 

### Liu hierarchy for children

1. Injury 
2. Measles 
3. Meningitis 
4. Malaria 
5. AIDS 
6. Diarrhea 
7. Acute Respiratory Infection (ARI) 
8. Possible pneumonia 
9. Possible diarrhea 
10. Unspecified 


## Implementation challenges

VA data is notoriously hard to analyze; survey instruments vary from place to place and over time, and the data 
collected varies considerably in quality and completeness. These issues complicate the application of EAVA to
VA data, despite the simplicity of the hierarchies outlined above. For instance, here are the criteria for
diarrhea in children according to @kalter:

> More frequent loose or liquid stools than usual AND more than 4 stools on the day with the most stools AND no blood in the stools 

OR

> More frequent loose or liquid stools than usual for more than 14 days AND no blood in stools

These criteria can only be checked if the relevant data are available; if there is no information about the duration of 
the child's diarrhea or number of stools, then the algorithm cannot justify identifying diarrhea as a cause of death or 
possible comorbidity.  Here is another example from the @kalter specification: a child decedent is deemed to
have suffered from underlying malnutrition if

> Limbs became very thin during the fatal illness OR had swollen legs or feet during the illness

AND

> One of these was the first symptom of the illness

While some of the VA datasets contain information about thin limbs or swollen appendages, we cannot use the 
available data to check this last condition. To mediate between these specific requirements and the 
shortcomings in our VA data, we have necessarily taken some liberties with our EAVA implementations. All of our 
code is [publicly available](https://github.com/databrew/eava), however, so anyone interested can inspect it 
to see exactly what has been done.

## Standardization of VA data

One of the major difficulties in analyzing VA data stems from the use of different survey instruments; these yield datasets
with wildly different names and numbers of variables. To mitigate this issue, we rely on the previously developed
open source package `vida` to translate each VA dataset into a standard `babel` format with standard variable names. 
Once this translation has been done, we can apply our EAVA implementations to the standardized data. A shortcoming 
of this approach is that newer survey instruments cannot yet be mapped to `babel` format. 

## InterVA

One of the most well-known algorithms for VA analysis is InterVA, an algorithm pioneered by Peter Byass and implemented
in various forms over the years (InterVA4, InterVA5, openVA versions). While InterVA has performed reasonably well on particular
datasets (add references!), it has rigid input requirements that prohibit its application to the VA data that we have.
In addition, InterVA is an opaque algorithm that depends on mysteriously tuned conditional probabilities; an InterVA user
cannot interrogate its cause of death findings, making its conclusions difficult to interpret. These were our
motivations for implementing open source expert algorithms: we needed algorithms that we could apply to a variety of VA datasets, 
and we wanted results that we could easily understand.


```{r}
# causes is a data.frame created by get_causes; 3rd column contains cause(s) of death determined by EAVA
# 1st entry is primary cause, others are possible comborbidities

clean_causes <- function( causes ){
  primary <- character()
  comorbidities <- character()
  for(i in 1:nrow(causes)){
    all <- unlist( strsplit( causes$causes[i], ", ", fixed = TRUE))
    if( length( all ) == 1 ){
      primary[i] <- all
      comorbidities[i] <- NA
    } else{
      primary[i] <- all[1]
      comorbidities[i] <- paste0( all[-1], collapse=", ")
    }
  }
  return( causes %>% 
            mutate( Primary=primary, Comorbidities = comorbidities ) %>% 
            select(-causes) %>% 
            rename("Age in days" = age_days, "Age group" = age_group) )
  
}

tally_causes <- function( cleaned_causes, age_group ){

  tallies <- cleaned_causes %>%
    filter( `Age group` == age_group) %>%
    group_by( Primary ) %>%
    tally( name="Count") 

  comorbid <-  cleaned_causes %>% 
    filter( `Age group` == age_group) %>%
    group_by(Primary) %>% 
    drop_na(Comorbidities) %>%
    summarize( Comorbidities = paste0( unique( unlist( strsplit( paste0(Comorbidities, collapse = ", "), ", ", fixed=TRUE))), collapse = ", "))

  return( tallies %>% left_join( comorbid, by="Primary" ) %>% arrange( desc(Count) ) )
}

print_causes <- function( causes, age_group="Child" ){
  
  tally_causes( clean_causes( causes ), age_group ) %>%
      rename( "Possible comorbidities" = Comorbidities) %>% 
      regulartable() %>% 
      # set_caption( paste0("Causes of death for ", ifelse(age_group=="Child", "children", "neonates"))) %>%
      theme_zebra() %>% 
      autofit()
  
}

diarrhea_table <- function( df ){
  # if( any( !is.na( df$diarrhea ) ) ){
  tab <- as.data.frame( table( df$diarrhea, useNA = "ifany" ) )
  colnames( tab ) <- c("Response", "Count")
  tab <- tab %>%
    mutate( Response = as.character(Response)) %>%
    replace_na( list(Response = "NA")) %>%
    mutate( Response = recode_factor(.x = Response,
                                     "dk" = "Don't know",
                                     "no" = "No",
                                     "yes" = "Yes",
                                     "Sum" = "Total")) %>%
    mutate( Percentage = paste0( round(100*Count / sum(Count)), "%" ) ) 
  regulartable(tab) %>% align(j=2:3, align="right", part="all")
  # } 
}

age_group_table <- function( causes ){
  tab <- as.data.frame( table( causes$age_group ) )
  colnames( tab ) <- c("Age group", "Count")
  regulartable(tab) 
}

```

# Results 

For each country, the name of each subsection is the name of the analyzed file (before `babel` translation).

## Mali data

### CHAMPS_Verbal_Autopsy_Form_results

Overall diarrhea incidence, based on answers to the standardized questions:

> Did the child have more frequent loose or liquid stools than usual? / Did the child have diarrhea?

```{r}
df <- read_csv("data/mali_babel/CHAMPS_Verbal_Autopsy_Form_results_babel.csv", show_col_types = FALSE)
diarrhea_table( df )
```

#### Age group breakdown 

```{r}
causes <- get_causes(df, format="champs2", algo="kalter")
all_causes <- causes %>%
  mutate( country = "Mali",
          filename = "data/mali_babel/CHAMPS_Verbal_Autopsy_Form_results_babel.csv")
age_group_table( causes )
```

#### Available standardized variables

This table is gross &mdash; it's just here as a reminder to discuss better ways to summarize available data per file.

```{r}
gv <- ls()[grep("_available", ls())]
available <- data.frame( )
for(i in 1:length(gv)){
  available <- bind_rows( available, 
                      data.frame( "Variable" = unlist(strsplit(gv[i],"_available")), Available=get(gv[i])) )
}
regulartable( available ) %>% autofit()

```

#### Kalter's EAVA for children

```{r}
print_causes( causes )
```

#### Kalter's EAVA for neonates

```{r}
print_causes( causes, age_group = "Neonate" )
```

#### Liu's EAVA for children

```{r}
causes <- get_causes(df, format="champs2", algo="liu")
all_causes <- rbind( all_causes, 
                     causes %>% 
                       mutate( country = "Mali", 
                               filename = "data/mali_babel/CHAMPS_Verbal_Autopsy_Form_results_babel.csv") )
print_causes( causes )
```

### CHAMPS_Verbal_Autopsy_Form_v2_04_1_results

Overall diarrhea incidence, based on answers to the standardized questions:

> Did the child have more frequent loose or liquid stools than usual? / Did the child have diarrhea?

```{r}
df <- read_csv("data/mali_babel/CHAMPS_Verbal_Autopsy_Form_v2_04_1_results_babel.csv", show_col_types = FALSE)
diarrhea_table( df )
```

#### Age group breakdown 

```{r}
causes <- get_causes(df, format="champs2", algo="kalter")
all_causes <- rbind( all_causes, 
                     causes %>% 
                       mutate( country = "Mali", 
                               filename = "data/mali_babel/CHAMPS_Verbal_Autopsy_Form_v2_04_1_results_babel.csv") )
age_group_table( causes )
```

#### Kalter's EAVA for children

```{r}
print_causes( causes )
```

#### Kalter's EAVA for neonates

```{r}
print_causes( causes, age_group = "Neonate" )
```

#### Liu's EAVA for children

```{r}
causes <- get_causes(df, format="champs2", algo="liu")
all_causes <- rbind( all_causes, 
                     causes %>% 
                       mutate( country = "Mali", 
                               filename = "data/mali_babel/CHAMPS_Verbal_Autopsy_Form_v2_04_1_results_babel.csv") )
print_causes( causes )
```

### CHAMPS_Verbal_Autopsy_Form_v2_04_results

Overall diarrhea incidence, based on answers to the standardized questions:

> Did the child have more frequent loose or liquid stools than usual? / Did the child have diarrhea?

```{r}
df <- read_csv("data/mali_babel/CHAMPS_Verbal_Autopsy_Form_v2_04_results_babel.csv", show_col_types = FALSE)
diarrhea_table( df )
```

#### Age group breakdown 

```{r}
causes <- get_causes(df, format="champs2", algo="kalter")
all_causes <- rbind( all_causes, 
                     causes %>% 
                       mutate( country = "Mali", 
                               filename = "data/mali_babel/CHAMPS_Verbal_Autopsy_Form_v2_04_results_babel.csv") )
age_group_table( causes )
```

#### Kalter's EAVA for children

```{r}
print_causes( causes )
```

#### Kalter's EAVA for neonates

```{r}
print_causes( causes, age_group = "Neonate" )
```

#### Liu's EAVA for children

```{r}
causes <- get_causes(df, format="champs2", algo="liu")
all_causes <- rbind( all_causes, 
                     causes %>% 
                       mutate( country = "Mali", 
                               filename = "data/mali_babel/CHAMPS_Verbal_Autopsy_Form_v2_04_results_babel.csv") )
print_causes( causes )
```

### VA_Death_of_a_child_under_4_weeks_to_59_months_results_21MAR2018

Overall diarrhea incidence, based on answers to the standardized questions:

> Did the child have more frequent loose or liquid stools than usual? / Did the child have diarrhea?

```{r}
df <- read_csv("data/mali_babel/VA_Death_of_a_child_under_4_weeks_to_59_months_results_21MAR2018_babel.csv", show_col_types = FALSE)
diarrhea_table( df )
```

#### Age group breakdown 

```{r}
causes <- get_causes(df, format="champs1", algo="kalter")
all_causes <- rbind( all_causes, 
                     causes %>% 
                       mutate( country = "Mali", 
                               filename = "data/mali_babel/VA_Death_of_a_child_under_4_weeks_to_59_months_results_21MAR2018_babel.csv") )
age_group_table( causes )
```

#### Kalter's EAVA for children

```{r}
print_causes( causes )
```

#### Kalter's EAVA for neonates

```{r}
print_causes( causes, age_group = "Neonate" )
```

#### Liu's EAVA for children

```{r}
causes <- get_causes(df, format="champs1", algo="liu")
all_causes <- rbind( all_causes, 
                     causes %>% 
                       mutate( country = "Mali", 
                               filename = "data/mali_babel/VA_Death_of_a_child_under_4_weeks_to_59_months_results_21MAR2018_babel.csv") )
print_causes( causes )
```

## Gambia results

### bn_hdss/VA_Child_WHO2016

Overall diarrhea incidence, based on answers to the standardized questions:

> Did the child have more frequent loose or liquid stools than usual? / Did the child have diarrhea?

```{r}
df <- read_csv("data/gambia_babel/bn_hdss/VA_Child_WHO2016_babel.csv", show_col_types = FALSE)
diarrhea_table( df )
```

#### Age group breakdown 

```{r}
causes <- get_causes(df, format="c16", algo="kalter")
all_causes <- rbind( all_causes, 
                     causes %>% 
                       mutate( country = "Gambia", 
                               filename = "data/gambia_babel/bn_hdss/VA_Child_WHO2016_babel.csv") )
age_group_table( causes )
```

#### Kalter's EAVA for children

```{r}
print_causes( causes )
```

#### Kalter's EAVA for neonates

```{r}
print_causes( causes, age_group = "Neonate" )
```

#### Liu's EAVA for children

```{r}
causes <- get_causes(df, format="c16", algo="liu")
all_causes <- rbind( all_causes, 
                     causes %>% 
                       mutate( country = "Gambia", 
                               filename = "data/gambia_babel/bn_hdss/VA_Child_WHO2016_babel.csv") )
print_causes( causes )
```

### bn_hdss/VA_ChildForm_WHO2012

Overall diarrhea incidence, based on answers to the standardized questions:

> Did the child have more frequent loose or liquid stools than usual? / Did the child have diarrhea?

```{r}
df <- read_csv("data/gambia_babel/bn_hdss/VA_ChildForm_WHO2012_babel.csv", show_col_types = FALSE)
diarrhea_table( df )
```

#### Age group breakdown 

```{r}
causes <- get_causes(df, format="c12_16", algo="kalter")
all_causes <- rbind( all_causes, 
                     causes %>% 
                       mutate( country = "Gambia", 
                               filename = "data/gambia_babel/bn_hdss/VA_ChildForm_WHO2012_babel.csv") )
age_group_table( causes )
```

#### Kalter's EAVA for children

```{r}
print_causes( causes )
```

#### Kalter's EAVA for neonates

```{r}
print_causes( causes, age_group = "Neonate" )
```

#### Liu's EAVA for children

```{r}
causes <- get_causes(df, format="c12_16", algo="liu")
all_causes <- rbind( all_causes, 
                     causes %>% 
                       mutate( country = "Gambia", 
                               filename = "data/gambia_babel/bn_hdss/VA_ChildForm_WHO2012_babel.csv") )
print_causes( causes )
```

### bs_hdss/VA_Child_Indepth_2008-2012_clean duration

Overall diarrhea incidence, based on answers to the standardized questions:

> Did the child have more frequent loose or liquid stools than usual? / Did the child have diarrhea?

```{r}
df <- read_csv("data/gambia_babel/bs_hdss/VA_Child_Indepth_2008-2012_clean duration_babel.csv", show_col_types = FALSE)
diarrhea_table( df )
```

#### Age group breakdown 

```{r}
causes <- get_causes(df, format="c08_12", algo="kalter")
all_causes <- rbind( all_causes, 
                     causes %>% 
                       mutate( country = "Gambia", 
                               filename = "data/gambia_babel/bs_hdss/VA_Child_Indepth_2008-2012_clean duration_babel.csv") )
age_group_table( causes )
```

#### Kalter's EAVA for children

```{r}
print_causes( causes )
```

#### Kalter's EAVA for neonates

```{r}
print_causes( causes, age_group = "Neonate" )
```

#### Liu's EAVA for children

```{r}
causes <- get_causes(df, format="c08_12", algo="liu")
all_causes <- rbind( all_causes, 
                     causes %>% 
                       mutate( country = "Gambia", 
                               filename = "data/gambia_babel/bs_hdss/VA_Child_Indepth_2008-2012_clean duration_babel.csv") )
print_causes( causes )
```

### bs_hdss/VA_Child_Indepth_2008-2012

Overall diarrhea incidence, based on answers to the standardized questions:

> Did the child have more frequent loose or liquid stools than usual? / Did the child have diarrhea?

```{r}
df <- read_csv("data/gambia_babel/bs_hdss/VA_Child_Indepth_2008-2012_babel.csv", show_col_types = FALSE)
diarrhea_table( df )
```

#### Age group breakdown 

```{r}
causes <- get_causes(df, format="c08_12", algo="kalter")
all_causes <- rbind( all_causes, 
                     causes %>% 
                       mutate( country = "Gambia", 
                               filename = "data/gambia_babel/bs_hdss/VA_Child_Indepth_2008-2012_babel.csv") )
age_group_table( causes )
```

#### Kalter's EAVA for children

```{r}
print_causes( causes )
```

#### Kalter's EAVA for neonates

```{r}
print_causes( causes, age_group = "Neonate" )
```

#### Liu's EAVA for children

```{r}
causes <- get_causes(df, format="c08_12", algo="liu")
all_causes <- rbind( all_causes, 
                     causes %>% 
                       mutate( country = "Gambia", 
                               filename = "data/gambia_babel/bs_hdss/VA_Child_Indepth_2008-2012_babel.csv") )
print_causes( causes )
```

### bs_hdss/VA_Child_WHO2012-mid 2016

Overall diarrhea incidence, based on answers to the standardized questions:

> Did the child have more frequent loose or liquid stools than usual? / Did the child have diarrhea?

```{r}
df <- read_csv("data/gambia_babel/bs_hdss/VA_Child_WHO2012-mid 2016_babel.csv", show_col_types = FALSE)
diarrhea_table( df )
```

#### Age group breakdown 

```{r}
causes <- get_causes(df, format="c12_16", algo="kalter")
all_causes <- rbind( all_causes, 
                     causes %>% 
                       mutate( country = "Gambia", 
                               filename = "data/gambia_babel/bs_hdss/VA_Child_WHO2012-mid 2016_babel.csv") )
age_group_table( causes )
```

#### Kalter's EAVA for children

```{r}
print_causes( causes )
```

#### Kalter's EAVA for neonates

```{r}
print_causes( causes, age_group = "Neonate" )
```

#### Liu's EAVA for children

```{r}
causes <- get_causes(df, format="c12_16", algo="liu")
all_causes <- rbind( all_causes, 
                     causes %>% 
                       mutate( country = "Gambia", 
                               filename = "data/gambia_babel/bs_hdss/VA_Child_WHO2012-mid 2016_babel.csv") )
print_causes( causes )
```

### bs_hdss/VA_Child_WHO2016

Overall diarrhea incidence, based on answers to the standardized questions:

> Did the child have more frequent loose or liquid stools than usual? / Did the child have diarrhea?

```{r}
df <- read_csv("data/gambia_babel/bs_hdss/VA_Child_WHO2016_babel.csv", show_col_types = FALSE)
diarrhea_table( df )
```

#### Age group breakdown 

```{r}
causes <- get_causes(df, format="c16", algo="kalter")
all_causes <- rbind( all_causes, 
                     causes %>% 
                       mutate( country = "Gambia", 
                               filename = "data/gambia_babel/bs_hdss/VA_Child_WHO2016_babel.csv") )
age_group_table( causes )
```

#### Kalter's EAVA for children

```{r}
print_causes( causes )
```

#### Kalter's EAVA for neonates

```{r}
print_causes( causes, age_group = "Neonate" )
```

#### Liu's EAVA for children

```{r}
causes <- get_causes(df, format="c16", algo="liu")
all_causes <- rbind( all_causes, 
                     causes %>% 
                       mutate( country = "Gambia", 
                               filename = "data/gambia_babel/bs_hdss/VA_Child_WHO2016_babel.csv") )
print_causes( causes )
```

## Overall results

### Mali

#### Age group breakdown 

```{r}
age_group_table( all_causes %>% filter( country == "Mali") )
```

#### All causes

```{r}
print_causes( all_causes %>% filter( country == "Mali") )
```

### Gambia

#### Age group breakdown 

```{r}
age_group_table( all_causes %>% filter( country == "Gambia") )
```

#### All causes

```{r}
print_causes( all_causes  %>% filter( country == "Gambia") )
```

### Mali and Gambia 

#### Age group breakdown 

```{r}
age_group_table( all_causes )
```

#### All causes

```{r}
print_causes( all_causes )
```


# Implementations 

## Kalter's EAVA for children

```{r echo=TRUE, eval=FALSE}

# Injury -----

# Kalter :

# Suffered from motor vehicle accident, fall, drowning, poisoning, venomous bite or sting, burn, violence or other injury
# AND 
# (Died 1 day or less after the injury AND the illness lasted 1 day or less) OR 
# (Injury and No other VA diagnosis (except malnutrition allowed)) OR 
# (Injury that was the first illness sign/symptom AND had VA other infection or fever))

injury <- function( responses ){
  questions <- c("injury_accident", "road_accident", "venomous_animal", "animal", "burn",
                 "drown", "injury_fall", "other_injury", "poisoning", "assault")
  questions <- questions[ which( questions %in% names(responses)) ]
  answers <- responses %>% select( all_of(questions) ) 
  if(any(!is.na(answers))) {
    if (any(answers == "yes", na.rm = TRUE)) {
      return(TRUE)
    } else{
      return(FALSE)
    }
  } else{ 
    return(FALSE)
  }
}

# CHILD causes of death according to Kalter -----

# AIDS (Kalter) -----

# (Swelling in the armpits OR a whitish rash inside the mouth/on the tongue)
# AND
# 3 or more of the following 6 signs: 
# (limbs became very thin, protruding belly, more frequent loose/liquid stools than usual for more than 30 days, 
# fever or a skin rash for more than 30 days, fast breathing, chest indrawing)

aids_kalter <- function( armpits, rash_mouth, thin_limbs, protruding_abdomen, fast_breathing, 
                  chest_indrawing, days_diarrhea, days_fever, days_rash ){
  if( (armpits | rash_mouth) ){
    signs <- thin_limbs + protruding_abdomen + fast_breathing + chest_indrawing + (days_diarrhea > 30) + (days_fever > 30) + (days_rash > 30) 
    return( signs >= 3 )
  } else{
    return(FALSE)
  }
}

# Malnutrition (underlying) (Kalter) -----

# Limbs became very thin during the fatal illness OR had swollen legs or feet during the illness
# AND
# One of these was the first symptom of the illness

underlying_malnutrition_kalter <- function( thin_limbs, swollen_legs_feet ){
  return( thin_limbs | swollen_legs_feet )
}

# Measles (Kalter) -----

# Child's age greater than or equal to 120 days 
# AND 
# rash for 3 or more days 
# AND
# fever for 3 or more days 
# AND 
# the rash started on the face

measles_kalter <- function( age_days, days_rash, days_fever ){
  return( (age_days > 120) & (days_rash >= 3) & (days_fever >= 3) )
}

# Meningitis (Kalter) -----

# Fever AND (stiff neck OR bulging fontanelle)

meningitis_kalter <- function( fever, stiff_neck, bulging_fontanelle ){
  return( fever & ( stiff_neck | bulging_fontanelle ) )
}

# Dysentery (Kalter) -----

# More frequent loose or liquid stools than usual AND more than 4 stools on the day with the most stools AND blood in the stools
# OR
# More frequent loose or liquid stools than usual for more than 14 days AND blood in the stools

dysentery_kalter <- function( days_diarrhea, bloody_stool ){
  return( (days_diarrhea > 14) & bloody_stool )
}

# Diarrhea (Kalter) -----

# More frequent loose or liquid stools than usual AND more than 4 stools on the day with the most stools AND No blood in the stools
# OR
# More frequent loose or liquid stools than usual for more than 14 days AND No blood in stools

diarrhea_kalter <- function( diarrhea, days_diarrhea, bloody_stool, number_stools ){
  return( (diarrhea & (number_stools > 4) & !bloody_stool ) | 
          ( (days_diarrhea > 14) & !bloody_stool ) )
}

# Pertussis (Kalter) -----

# Cough more than 14 days AND 
# (severe cough OR vomited after coughing OR stridor)

pertussis_kalter <- function( days_cough, cough_severe_available, cough_severe, cough_vomit_available, cough_vomit ){
  if( cough_severe_available & cough_vomit_available ){
    return( (days_cough > 14) & (cough_severe | cough_vomit) )
  } else if( cough_severe_available & !cough_vomit_available){
    return( (days_cough > 14) & cough_severe ) 
  } else if( !cough_severe_available & cough_vomit_available){
    return( (days_cough > 14) & cough_vomit ) 
  } else{ 
    return( (days_cough > 14) )
  }
}

# Pneumonia (Kalter) -----

# (Cough more than 2 days OR difficult breathing more than 2 days)
# AND
# (Fast breathing more than 2 days OR chest indrawing OR grunting)

pneumonia_kalter <- function( days_cough, days_difficulty_breathing, days_fast_breathing, chest_indrawing, noisy_breathing ){
  return( ( (days_cough > 2) | (days_difficulty_breathing > 2)) & 
            ( (days_fast_breathing > 2) | chest_indrawing | noisy_breathing ) )
}

# Malaria (Kalter) -----

# Fever that continued till death AND was on and off in character AND No stiff neck AND 
# No bulging fontanelle AND (pallor OR difficult breathing OR convulsions OR unconscious till death)
# OR
# Fever that continued till death AND was severe fever AND No stiff neck AND 
# No bulging fontanelle AND (pallor OR convulsions OR unconscious till death)

malaria_kalter <- function( fever_continue, fever_on_off, fever_severe, stiff_neck, bulging_fontanelle, 
                     difficulty_breathing, convulsions, unconscious ) {
  return( fever_continue & (fever_on_off | fever_severe) & !stiff_neck & !bulging_fontanelle & 
            (difficulty_breathing | convulsions | unconscious) )
}

# Possible dysentery (Kalter) -----

# More frequent loose or liquid stools than usual AND 
# (fever OR convulsions OR unconscious up till death) AND 
# blood in the stools AND 
# No VA dysentery

possible_dysentery_kalter <- function( diarrhea, fever, convulsions, unconscious, bloody_stool ){
  return( diarrhea & (fever | convulsions | unconscious) & bloody_stool )
}

# Possible diarrhea (Kalter) -----

# More frequent loose or liquid stools than usual AND 
# (fever OR convulsions OR unconscious up till death) AND 
# No blood in the stools AND 
# No VA diarrhea

possible_diarrhea_kalter <- function( diarrhea, fever, convulsions, unconscious, bloody_stool ){
  return( diarrhea & (fever | convulsions | unconscious) & !bloody_stool )
}

# Possible pneumonia (Kalter) -----

# (Cough or difficult breathing) OR (Fast breathing AND (chest indrawing OR stridor OR grunting OR wheezing))
# AND
# (Severe cough OR post–tussive vomiting OR fast breathing OR chest indrawing OR grunting OR stridor OR wheezing OR 
# fever OR convulsions OR unconscious up till death) 
# AND 
# No VA Pertussis AND No VA pneumonia

possible_pneumonia_kalter <- function( cough, difficulty_breathing, fast_breathing, chest_indrawing, noisy_breathing, 
                                cough_severe, cough_vomit, fever, convulsions, unconscious ){
  return( ( (cough | difficulty_breathing) | ( fast_breathing & (chest_indrawing | noisy_breathing) ) ) & 
            ( cough_severe | cough_vomit | fast_breathing | chest_indrawing | noisy_breathing | fever | convulsions | unconscious ) )
}

# Hemorrhagic fever (Kalter) -----

# Fever AND (bled from anywhere OR had areas of the skin that turned black)

hemorrhagic_fever_kalter <- function( fever, bled_anywhere, skin_black ){
  return( fever & (bled_anywhere | skin_black) )
}

# Other infection (Kalter) -----

# Fever AND 
# (rash on trunk, abdomen or everywhere OR convulsions OR unconscious up till death)

other_infection_kalter <- function( fever, rash_trunk, convulsions, unconscious ){
  return( fever & (rash_trunk | convulsions | unconscious) )
}

# Possible malaria (Kalter) -----

# Fever AND No other VA infectious causes of death

possible_malaria_kalter <- function( fever ){
  return( fever )
}

# Malnutrition (Kalter) -----

# Limbs became very thin during the fatal illness OR 
# had swollen legs or feet during the illness

malnutrition_kalter <- function( thin_limbs, swollen_legs_feet ){
  return( thin_limbs | swollen_legs_feet )
}

```

## Kalter's EAVA for neonates

```{r echo=TRUE, eval=FALSE}

# NEONATE causes of death according to Kalter

# Neonatal tetanus (Kalter) -----

# (Age 3–27 days at death AND convulsions or spasms)
# AND EITHER 
# ((Able to suckle normally during the first day of life and stopped being able to suckle) 
# OR 
# (cried within 5 minutes after birth and stopped being able to cry))

neonatal_tetanus_kalter <- function( age_days, convulsions, suckle_feed, stop_suckle, 
                                     cry_after_birth, baby_cry_after_birth, stop_ability_to_cry ){
  return( ( (age_days >=3) & (age_days <= 27) & convulsions ) & 
          ( ( suckle_feed & stop_suckle ) | ( (cry_after_birth | baby_cry_after_birth) & stop_ability_to_cry ) ) )
}

# Congenital malformation (Kalter) -----

# Gross malformation present at birth

malformation_kalter <- function( malformation ){
  return( malformation )
}

# Birth asphyxia (Kalter) -----

# Neonatal respiratory depression: (Did not cry within 5 minutes after birth OR did not breathe immediately after birth)
# AND
# Neonatal encephalopathy: (Not able to suckle normally in the first day of life OR convulsions/spasms OR lethargy) OR 0 days old at death

asphyxia_kalter <- function( age_days, cry_after_birth, baby_cry_after_birth, baby_breathe_after_birth, 
                             suckle_feed, convulsions, lethargic ){
  return( ( !(cry_after_birth | baby_cry_after_birth) | !baby_breathe_after_birth ) &
          ( !suckle_feed  | convulsions | lethargic | (age_days==0) ) )
}

# Birth injury (Kalter) -----

# Bruises or signs of injury on the body at birth

birth_injury_kalter <- function( bruises_injury, sign_injury_broken_bones ){
  return( bruises_injury | sign_injury_broken_bones )
}

# Preterm delivery with respiratory distress syndrome (Kalter) -----

# Pregnancy duration less than 9 months
# AND
# (Fast breathing starting on day 0 AND no fever AND no cold to touch)

preterm_respiratory_distress_kalter <- function( months_pregnancy, fast_breathing, fever, cold_touch ){
  return( ( months_pregnancy < 9 ) & ( fast_breathing & !fever & !cold_touch ) )
}

# Neonatal meningitis (Kalter) -----

# Fever AND (bulging fontanelle OR convulsions) AND (lethargic OR unresponsive/unconscious)

neonatal_meningitis_kalter <- function( fever, bulging_fontanelle, convulsions, lethargic, unconscious ){
  return( fever & (bulging_fontanelle | convulsions) & (lethargic | unconscious) )
}

# Neonatal diarrhea (Kalter) -----

# More frequent loose or liquid stools than usual AND more than 4 stools on the day the diarrhea was most frequent

neonatal_diarrhea_kalter <- function( diarrhea, number_stools ){
  return( diarrhea & (number_stools > 4) )
}

# Neonatal pneumonia (Kalter) -----

# (Fast breathing lasting 1 day or more OR difficult breathing lasting 1 day or more and lasting until death)
# AND
# 2 or more of the following 3 signs: (chest indrawing, grunting, never cried OR stopped crying)

neonatal_pneumonia_kalter <- function( days_difficulty_breathing, days_fast_breathing, chest_indrawing, 
                                       noisy_breathing, baby_cry, stop_ability_to_cry ){
  return( ( ( days_fast_breathing >= 1 ) | ( days_difficulty_breathing >= 1) ) &
          ( chest_indrawing + noisy_breathing + (!baby_cry | stop_ability_to_cry) > 2 ) )
}

# Neonatal possible diarrhea (Kalter) -----

#  More frequent loose or liquid stools than usual AND VA sepsis (see below) AND No VA diarrhea

neonatal_possible_diarrhea_kalter <- function( diarrhea, fever, cold_touch, suckle_feed, stop_suckle, convulsions, vomit, 
                                               stop_ability_to_cry, lethargic, unconscious, chest_indrawing, noisy_breathing){
  return( diarrhea & sepsis_kalter(fever, cold_touch, suckle_feed, stop_suckle, convulsions, vomit, 
                                   stop_ability_to_cry, lethargic, unconscious, chest_indrawing, noisy_breathing) )
}

# Neonatal possible pneumonia (Kalter) -----

# Difficult breathing AND VA sepsis AND No VA pneumonia

neonatal_possible_pneumonia_kalter <- function( difficulty_breathing, fever, cold_touch, suckle_feed, stop_suckle, convulsions, vomit, 
                                                stop_ability_to_cry, lethargic, unconscious, chest_indrawing, noisy_breathing){
  return( difficulty_breathing & sepsis_kalter(fever, cold_touch, suckle_feed, stop_suckle, convulsions, vomit, 
                                               stop_ability_to_cry, lethargic, unconscious, chest_indrawing, noisy_breathing) )
}

# Sepsis (Kalter) -----

# Fever OR cold to touch OR 2 or more of the following 7 signs: 
# (fever OR cold to touch, did not suckle normally on the first day of life OR stopped suckling, convulsions, 
# vomited everything, stopped crying, lethargic OR unconscious, chest indrawing OR grunting)

sepsis_kalter <- function( fever, cold_touch, suckle_feed, stop_suckle, convulsions, vomit, 
                           stop_ability_to_cry, lethargic, unconscious, chest_indrawing, noisy_breathing){
  return( (fever | cold_touch) |
          ( (fever | cold_touch) + (!suckle_feed | stop_suckle) + convulsions + vomit + stop_ability_to_cry + 
              (lethargic | unconscious) + (chest_indrawing | noisy_breathing) > 2 ) )
}

# Neonatal jaundice (Kalter) -----

# Yellow skin or yellow eyes 
# AND 
# (stopped being able to suckle normally OR lethargic OR unresponsive/unconscious) 
# AND 
# No fever or hypothermia

neonatal_jaundice_kalter <- function( jaundice, stop_suckle, lethargic, unconscious, fever ){
  return( jaundice & ( stop_suckle | lethargic | unconscious ) & !fever )
}

# Neonatal hemorrhagic syndrome (Kalter) -----

# Bleeding from anywhere AND No fever or cold to touch

neonatal_hemorrhagic_kalter <- function( bled_anywhere, fever, cold_touch ){
  return( bled_anywhere & (!fever | cold_touch) )
}

# Sudden unexplained death (Kalter) -----

# Died suddenly without appearing ill AND No illness signs or symptoms

neonatal_sudden_kalter <- function( die_suddenly, appear_healthy ){
  return( die_suddenly | appear_healthy )
}

# Preterm delivery (Kalter) -----

# Pregnancy duration less than 8 months

preterm_kalter <- function( months_pregnancy ){
  return( months_pregnancy < 8 )
}

```

## Liu's EAVA for children

```{r echo=TRUE, eval=FALSE}

# CHILD causes of death according to Liu -----

# Measles (Liu) -----

# a. Age at death>=6 months;
# AND
# b. Measles-type rash on body and face;
# AND
# c. Accompanied by fever;
# AND
# d. With at least 1 of the following specific symptoms: dry cough, or red or runny eyes.

measles_liu <- function( age_days, fever, measles_rash, cough, red_eyes ){
  return( (age_days > 180) & measles_rash & fever & (cough | red_eyes) )
}

# Meningitis (Liu) -----

# a. Age at death>=29 days;
# AND
# b. Fever;
# AND
# c. Convulsions;
# AND
# d1. Stiff neck;
# OR
# d2. Bulging fontanelle;
# AND
# e. With at least 1 of the following specific symptoms: unconscious.

meningitis_liu <- function( fever, convulsions, stiff_neck, bulging_fontanelle, unconscious ){
  return( fever & convulsions & ( stiff_neck | bulging_fontanelle ) & unconscious )
}

# Malaria (Liu) -----

# a. Age at death>=29 days;
# AND
# b. Fever;
# AND
# c. With at least 1 of the following specific symptoms: convulsions, difficult breathing, unresponsive, or pallor;

malaria_liu <- function( fever, difficulty_breathing, convulsions, unresponsive ) {
  return( fever & (difficulty_breathing | convulsions | unresponsive) )
}

# AIDS (Liu) -----

# a. Age at death>=29 days;
# AND
# b1. Jaundice;
# OR
# b2. Chronic diarrhea > 1 month;
# OR
# b3. Chronic fever > 1 month;
# OR
# b4. Wasting defined as having all the 4 following
# symptoms: paleness, hair color change, edema legs, dry
# scaly skin;
# OR
# b5. Cough or trouble breathing lasting 3-27 days with
# fever but not recent TB.

aids_liu <- function( jaundice, fever, days_diarrhea, days_fever, pale, hair_change, swollen_legs_feet, days_cough, days_difficulty_breathing, tb ){
  return( jaundice | (days_diarrhea >= 29) | (days_fever >= 29) | ( pale & hair_change & swollen_legs_feet ) | 
            ( ((days_cough >= 3) | (days_difficulty_breathing >= 3)) & fever & !tb ) )
}

# Diarrhea (Liu) -----

# a. Age at death>=29 days;
# AND
# b1. Frequent loose or liquid stools starting from 1 to 13
# days before death and continued until death;
# AND
# b2. With a peak number of 6 or more stools in 24 hours;
# AND
# b3. At least 2 of the 6 following specific symptoms were
# reported: weakness, dry mouth, sunken eyes, loose skin,
# depressed fontanels, and no or very little urine;

diarrhea_liu <- function( diarrhea, days_diarrhea, number_stools, sunken_eye, sunken_fontanelle ){
  return( (diarrhea & (days_diarrhea < 14) & (number_stools >= 6) & (sunken_eye & sunken_fontanelle)) | 
            (diarrhea & (days_diarrhea >= 14)) )
}

# Acute Respiratory Infection (Liu) -----

# a. Age at death>=29 days;
# AND
# b1. Had a cough that started at least 3 days before
# death;
# OR
# b2. Difficulty breathing that started at least 1 day before
# death;
# AND
# c. Had at least 2 of the following 6 specific symptoms: noisy
# breathing, grunting, wheezing, nostril flaring, or chest indrawing.

ari_liu <- function( days_cough, difficulty_breathing, noisy_breathing, chest_indrawing, flaring_nostrils ){
  return( ((days_cough >= 3) | difficulty_breathing ) & ( noisy_breathing + chest_indrawing + flaring_nostrils >= 2 ) )
}

# Possible pneumonia (Liu) -----

# a. Age at death>=29 days;
# AND
# b. Had at least 2 of the following signs of serious infection:
#   difficult breathing, chest indrawing, convulsions, and fever;
# AND
# c. Cough or difficult breathing.

possible_pneumonia_liu <- function( cough, difficulty_breathing, chest_indrawing, fever, convulsions ){
  return( ( difficulty_breathing + chest_indrawing + convulsions + fever >= 2 ) & (cough | difficulty_breathing) )
}

# Possible diarrhea (Liu) -----

# a. Age at death>=29 days ;
# AND
# b. The child had 2+ signs of serious infection: difficult
# breathing, chest indrawing, convulsions, and fever;
# AND
# c. Diarrhea.

possible_diarrhea_liu <- function( diarrhea, difficulty_breathing, chest_indrawing, fever, convulsions ){
  return( ( difficulty_breathing + chest_indrawing + convulsions + fever >= 2 ) & diarrhea )
}

```

# References
